---
title: "Unit 2 In Class Discussion"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
code-annotations: hover
bibliography: QMLS_Bibliography.bib
csl: evolution.csl
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)

```

## Miscellaneous questions

- Customizing plots
- Excel vs. csv vs. some other format
- Writing files from R
- What does the pipe do?
- When to use native pipe (`|>`) vs. magrittr pipe (`%>%`)?


## `tibble`s and `data.frame`s

- When are they the same?
- When are they different?
- Which one should I use?


## Pivoting

- Longer
    - Multiple columns into rows
    - We use 99% of the time
- Wider
    - Rows into multiple columns
    - We use 1% of the time


## Pivoting to long format

```{r}
#| echo: true

set.seed(423747)
DD <- tibble(ID = 1:10,
             Column_1 = runif(10),
             Column_2 = rnorm(10, mean = 10))
DD
```


## Pivoting to long format

```{r}
#| echo: true

DD_long <- DD |> 
  pivot_longer(cols = -ID,
               names_to = "old_column",
               values_to = "random_number")
DD_long
```


## Plotting long data

```{r}
#| echo: true
#| output-location: slide

ggplot(DD_long, aes(x = ID,
                    y = random_number,
                    color = old_column)) +
  geom_point(size = 3) +
  theme_bw()
```


## Pivoting to wide format

```{r}
#| echo: true
DD_long |> 
  pivot_wider(names_from = old_column,
              values_from = random_number)
```


## The pivot round trip

1. Pivot to long format
2. Do some calculations
3. Pivot to wide format


## Learning R commands


## Base R vs. tidyverse?

- What should I learn?
- When should I use which?


## Why do we learn multiple tools?

Which tool is the best for what situation does not have a single answer


## When will we start learning statistics?

