---
title: "Unit 3 In Class Discussion"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
code-annotations: hover
bibliography: QMLS_Bibliography.bib
csl: evolution.csl
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
```


## From anonnymous questions

- Can you continuously edit/add to an excel sheet and have it load into a project with the newest version?

- Can you explain how to write relative paths? I tried ./PS_02/filename.xls and it wouldn't render. I tried other equivalent versions and it also wouldn't render. The only thing that worked was the exact path. I fear I may be misunderstanding relative paths. 


## You should have a question.

- You can't understand everything
- Something you are curious about

Spend some time thinking before you answer with your question.

- Take notes while you are watching
- Review the slides / videos


## How to use the problem set keys

Compare your answers and study alternate approaches


## Cheatsheets

- [Posit's tidyverse cheatsheets](https://posit.co/resources/cheatsheets/)
- [RStudio's repository](https://github.com/rstudio/cheatsheets)
    - [Base R](https://github.com/rstudio/cheatsheets/blob/main/base-r.pdf)


## Miscellaneous questions

- Fixing up plots, what goes where
- When am I using base vs. tidyverse? Does it matter?
- Understanding R's error messages


## For the end of class


## Pasting text together

- `paste()` and `paste0()`
    - Use inside `mutate()`
- `unite()`
    - Use in a pipe

```{r}
#| echo: true

(DD <- tibble(Col1 = letters[1:2],
              Col2 = 1:2))

```


## `paste()` and `paste0()`

```{r}
#| echo: true

DD |> 
  mutate(Paste = paste(Col1, Col2),
         Paste2 = paste(Col1, Col2, sep = "-"),
         Paste0 = paste0(Col1, Col2))
```


## `unite()`

By default the old columns are dropped (`remove = TRUE`)

```{r}
#| echo: true
DD |> unite(col = "Unite", Col1, Col2)
DD |> unite(col = "Unite", Col1, Col2, sep = "-")
```


```{r}
#| echo: true
DD |> unite(col = "Unite", Col1, Col2, remove = FALSE)
```


## Boxplots: What do all the parts mean?

```{r}
#| echo: true
#| output-location: slide

set.seed(3423644)
DD <- tibble(y = c(rnorm(19, mean = 10, sd = 5), 25))

ggplot(DD, aes(y)) +
  geom_boxplot() +
  coord_flip()
```


## Boxplot recipe

1. Median = center line
1. Box = IQR (25th to 75th percentiles)
1. Whiskers at 1.5 x 25th and 75th percentiles
1. Dots for anything outside 1.5 x 25th and 75th percentiles


## `separate()` vs. stringr functions
