---
title: "From Probability to Likelihood"
subtitle: "Discrete Data"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
code-annotations: hover
bibliography: QMLS_Bibliography.bib
csl: evolution.csl
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(cowplot)
library(forcats)
library(latex2exp)
theme_set(theme_cowplot())
```


## Frequency Distribution of Heads | 6 Flips

```{r}
#| fig.height: 4.5

flips <- tibble(
  Ways = c(1, 6, 15, 20, 15, 6, 1),
  Flips = c("0 H", "1 H", "2 H",
            "3 H", "4 H", 
            "5 H", "6 H")) |> 
  mutate(Flips = fct_inorder(Flips))

flips |> ggplot(aes(x = Flips, y = Ways)) +
  geom_bar(stat = "Identity")
```


## Probability Distribution of Heads | 6 Flips

```{r}
#| echo: true

flips$Ways
flips$Ways / sum(flips$Ways)
```


## Probability Distribution of Heads | 6 Flips

```{r}
#| fig.height: 3.5

flip6 <- tibble(
  Ways = c(1, 6, 15, 20, 15, 6, 1),
  Flips = c("0 H", "1 H", "2 H",
            "3 H", "4 H", 
            "5 H", "6 H"),
  Prob = Ways / sum(Ways)) |> 
  mutate(Flips = fct_inorder(Flips))

flip6 |> 
  ggplot(aes(x = Flips, y = Prob)) +
  geom_bar(stat = "Identity") +
  labs(y = "Probability")
```

How can we go from a set of probabilities to likelihoods?


## Model Likelihood ($\mathcal{L}$)

Given data ($Y$) and *hypothesized* parameter values ($\Theta$):

$$\mathcal{L}\left(Y;\Theta\right) = \prod Pr\left[Y; \Theta\right]$$

- Where $Y$ is a set of observations ($Y_1$, $Y_2$, ..., $Y_n$)
- The model likelihood is the product ($\Pi$) of the observations' individual probabilities.


## Model Likelihood ($\mathcal{L}$)

$$\mathcal{L}\left(Y;\Theta\right) = \prod Pr\left[Y; \Theta\right]$$

- Evaluate the likelihood for different values of $\Theta$.
- Maximize $\mathcal{L}$ and you will have the maximum likelihood set of parameter estimates.


## Large and small numbers

The product of a *large* number of probabilities can result in some very *small* probabilities

- Computers don't handle really small numbers very well
- Lower limit to the smallest number a computer can keep track of: $`r .Machine$double.xmin`$ in R.

Think about computing the model likelihood with thousands or millions of observations.

```{r}
#| echo: true

0.09^150
```


## Maximize the natural log of $\mathcal{L}$

The log-likelihood is easier to deal with mathematically.

$$\mathcal{L}\left(Y;\Theta\right) = \prod Pr\left[Y; \Theta\right]$$

Log both sides of the equation:

$$\log \left( \mathcal{L}\left(Y;\Theta\right) \right) = \log \left( \prod Pr\left[Y; \Theta\right] \right)$$


## Mathematics of logarithms

$$\log \left( \mathcal{L}\left(Y;\Theta\right) \right) = \log \left( \prod Pr\left[Y; \Theta\right] \right)$$

The log of the products equals the sum of the logs:

$$\log \left( \mathcal{L}\left(Y;\Theta\right) \right) = \sum \log \left(Pr\left[Y; \Theta\right] \right)$$

Sum the log-likelihoods of the observations to get the model likelihood. 

_Note_: `log()` is _natural_ log.


## Likelihood of coin flips

You have 2 heads from 6 coin flips (i.e., 1 observation)

- What is the maximum likelihood estimate of the probability of heads $(\theta)$?
- $\theta$ can be any value from 0 to 1 (but we won't include 0 or 1).


## log-Likelihood for one value of $\theta$

Hypothesize that $\theta = 0.2$, observe 2 heads from 6 flips. What is the likelihood?

```{r}
#| echo: true

theta <- 0.2
pr <- dbinom(2, 6, prob = theta)
pr
log(pr)
```


## log-Likelihood for one value of $\theta$

Hypothesize that $\theta = 0.2$, observe 2 heads from 6 flips. What is the likelihood? 

\begin{align*}
\log \left( \mathcal{L}\left(Y;\Theta\right) \right) &= \sum \log \left(Pr\left[Y; \Theta\right] \right)\\

\log\left(\mathcal{L}\left( Y; 0.2 \right)\right) &=  \log\left( 0.24576 \right)
\end{align*}

Where $Y$ is 2 heads from 6 flips (2/6).


## Likelihood across the range of 0 to 1

Generate a range of possible values for $\theta$:

```{r}
#| echo: true

theta <- seq(0.001, 0.999, length = 200)
```

Calculate the probability for 2 heads for each value of $\theta$:

```{r}
#| echo: true

pr <- dbinom(2, 6, prob = theta)
```

Convert to log-likelihoods:

```{r}
#| echo: true
#| output-location: slide
log_lik <- log(pr)
log_liks <- tibble(theta, log_lik)

log_liks
```

```{r}
log_liks_comb <- log_liks |> 
  rename(log_lik_6 = log_lik)
```


## Likelihood across the range of 0 to 1

```{r}
log_liks |> 
  ggplot(aes(theta, log_lik)) +
  geom_line(size = 1.5) +
  labs(x = TeX("$\\theta$"), y = "log-Likelihood")
```


## Maximum likelihood

`which.max()` returns the index of the maximum value.

```{r}
#| echo: true

max(log_lik)
theta[which.max(log_lik)]
```

- 1/3 was not among the tested values for $\theta$.


## Maximum likelihood

```{r}
log_liks |> 
  ggplot(aes(theta, log_lik)) +
  geom_line(size = 1.5) +
  geom_point(aes(x = theta[which.max(log_lik)], y = max(log_lik)),
             color = "red", size = 3) +
  labs(x = TeX("$\\theta$"), y = "log-Likelihood")
```


## Likelihood of coin flips

You have 20 heads from 60 coin flips.

- What is the maximum likelihood estimate of the probability of heads ($\theta$)?
- $\theta$ can be any value from 0 to 1.


## Likelihood across the range of 0 to 1

```{r}
#| echo: true

theta <- seq(0.001, 0.999, length = 200)
pr <- dbinom(20, 60, prob = theta)
log_lik <- log(pr)
log_liks <- tibble(theta, log_lik)
log_liks
```

```{r}
log_liks_comb$log_lik_60 <- log_lik
```


## Likelihood across the range of 0 to 1

```{r}
log_liks |> 
  ggplot(aes(theta, log_lik)) +
  geom_line(size = 1.5) +
  labs(x = TeX("$\\theta$"), y = "log-Likelihood")
```


## Maximum likelihood

```{r}
#| echo: true

max(log_lik)
theta[which.max(log_lik)]
```


## Maximum likelihood

```{r}
log_liks |> 
  ggplot(aes(theta, log_lik)) +
  geom_line(size = 1.5) +
  geom_point(aes(x = theta[which.max(log_lik)], y = max(log_lik)),
             color = "red", size = 3) +
  labs(x = TeX("$\\theta$"), y = "log-Likelihood")
```


## Likelihood of coin flips

You have 20000 heads from 60000 coin flips.

- What is the maximum likelihood estimate of the probability of heads ($\theta$)?
- $\theta$ can be any value from 0 to 1.


## Likelihood across the range of 0 to 1

```{r}
#| echo: true

theta <- seq(0.001, 0.999, length = 200)
log_lik <- dbinom(20000, 60000, prob = theta, log = TRUE)
log_liks <- tibble(theta, log_lik)
```

*Note*: Use `log = TRUE` to return the log-probability directly from `dbinom()` to avoid problems with `log(`*very small number*`)`.

```{r}
log_liks_comb$log_lik_60000 <- log_lik
```


## Maximum likelihood

```{r}
log_liks |> 
  ggplot(aes(theta, log_lik)) +
  geom_line(size = 1.5) +
  geom_point(aes(x = theta[which.max(log_lik)], y = max(log_lik)),
             color = "red", size = 3) +
  labs(x = TeX("$\\theta$"), y = "log-Likelihood")
```


## Comparing likelihoods as sample size increases

```{r}
#| echo: true

head(log_liks_comb)
```


## Comparing likelihoods as sample size increases

```{r}
#| warning: false

log_liks_comb |> 
  pivot_longer(cols = -theta, names_to = "n", values_to = "log-Likelihood") |> 
  mutate(n = str_remove(n, "log_lik_") |> as.numeric() |> factor()) |> 
  ggplot(aes(x = theta, y = `log-Likelihood`, color = n)) +
  geom_line(size = 1.5) +
  labs(x = TeX("$\\theta$"), y = "log-Likelihood")
```


## Comparing likelihoods as sample size increases

```{r}
#| warning: false

log_liks_comb |> 
  pivot_longer(cols = -theta, names_to = "n", values_to = "log-Likelihood") |> 
  mutate(n = str_remove(n, "log_lik_") |> as.numeric() |> factor()) |> 
  ggplot(aes(x = theta, y = `log-Likelihood`, color = n)) +
  geom_line(size = 1.5) +
  labs(x = TeX("$\\theta$"), y = "log-Likelihood") +
  ylim(c(-200, 0))
```


## Describing discrete data: 2 heads / 6 flips

- Notice that the ML estimate was ~0.33
- The mean of a discrete set of trials is k / n: 2 / 6

```{r}
#| echo: true

2/6
mean(c(1, 1, 0, 0, 0, 0))
mean(c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE))
```
