---
title: "Model Synonyms"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
code-annotations: hover
bibliography: QMLS_Bibliography.bib
csl: evolution.csl
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
library(cowplot)
theme_set(theme_cowplot())

```

## Linear models

- General term for a linear relationship between an outcome (observed) variable and one or more predictors
- A functional relationship + "noise"
- Most common type of statistical model

$$\mbox{Outcome variable} \sim \mbox{Predictor variable(s)} + \mbox{Error}$$


## Horned Lizards

$$Horn Length = \theta_0 + \theta_1 Group$$

```{r}

set.seed(3575575)

Alive <- rnorm(n = 150, mean = 24.5, sd = 2.6)
Dead <- rnorm(n = 30, mean = 22.0, sd = 2.7)
Group <- c(rep("Alive", 150),
           rep("Dead", 30))

HL <- tibble(Horn_Length = c(Alive, Dead),
             Group = factor(Group))

```

This same model can be implemented via:

- t-test (`t.test()`)
- linear regression model (`lm()`)
- ANOVA `aov()` or `anova(lm())`

## t-test

```{r}
#| echo: true

t.test(Horn_Length ~ Group, data = HL, var.equal = TRUE)
```

## Linear regression model 

```{r}
#| echo: true

fm <- lm(Horn_Length ~ Group, data = HL)
summary(fm)
```

## ANOVA 

```{r}
#| echo: true

fm <- lm(Horn_Length ~ Group, data = HL)
anova(fm)

```

## Categorical variables as numbers

```{r}
#| echo: true

model.matrix(fm)[145:155,]

```

## Categorical variables coded as 0/1

```{r}
#| echo: true

HL <- HL |>
  mutate(Group_num = ifelse(Group=="Alive",0,1))

HL

```

## Categorical variables coded as 0/1

```{r}
#| echo: true

fm <- lm(Horn_Length ~ Group_num, data = HL)
summary(fm)


```


## Named tests

- `lm()` is always applicable for a simple linear model
- `t.test` and ANOVA functions 
    - apply to a subset of models
        - e.g., categorical predictor
    - are shorthand for a particular subset of results
        - e.g., print the sums of squares

